---
title: "Section 7: Statistical Inference"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

```{r}
library(tidyverse)
```

## Hypothesis testing

A device advertised as "90% effective" only works 24 out of 30 days.  Is this noteworthy?  

Let's imagine what a typical 30 days would look like if the advertisement was correct.

```{r}
# Flip a "90-10" coin 30 times
results <- sample(0:1, size = 30, replace = TRUE,
                  prob = c(0.1, 0.9))

results

sum(results)
mean(results)
```

Let's repeat this sampling procedure 10000 times to get a sampling distribution of "if the advertising is correct".

```{r}
proportions <- tibble(prop_correct = replicate(10000,
                         mean(sample(0:1, size = 30, replace = TRUE,
                         prob = c(0.1, 0.9)))
                         )
                      )
```

```{r}
ggplot(proportions, aes(x = prop_correct)) +
  geom_histogram(fill = "dodgerblue", color = "black", bins = 12) +
```

Now let's add in our observed 24/30 days.

```{r}
obs <- 24/30

ggplot(proportions, aes(x = prop_correct)) +
  geom_histogram(fill = "dodgerblue", color = "black", bins = 12) +
  geom_vline(xintercept = obs, color = "forestgreen",
             linewidth = 1.5)
```

The resamples are called the null distribution - they represent the null hypothesis being true.

The p-value is the proportion of resamples *as extreme* or *more extreme* than our observed value.

```{r}
pval <- mean(proportions$prop_correct <= 0.8)

pval
```

EX 7.1

In a sample of 300 births near power plants, 52% were girls? Is this noteworthy?  Perform a simulation experiment where the null hypothesis is that p = 0.5, meaning girls and boys are equally likely.

Then, compare your resampled values to the outcome with 300(0.52) = 156 girl births.


## Two-sample test

Let's return to the two-sample penguin example, where we are comparing the median weight of male and female Adelie penguins.

```{r}
penguins <- read_csv("../../data/penguins.csv") %>%
  relocate(flipper_length_mm, .after = species)

adelie <- penguins %>% filter(species == "Adelie")

new_data <- function(data, n) {
  sample_n(data, size = n, replace = TRUE)
}
```

The "big box" resampling corresponds to a null hypothesis that the median weights are the same for the two groups.

```{r}
median_diffs <- tibble(
  run = 1:10000,
  female = replicate(10000,
                    median(new_data(adelie, 73)$body_mass_g) ),
  male = replicate(10000,
                    median(new_data(adelie, 73)$body_mass_g) ),
  diff = female - male
)

median_diffs
```

```{r}
ggplot(median_diffs, aes(x = diff)) +
  geom_histogram(fill = "dodgerblue", color = "black") +
  geom_vline(xintercept = -600,
             color = "forestgreen", linewidth = 1.5)
```

Two-sided p-value calculation:

```{r}
# "as or more extreme" than 600 on BOTH sides
mean(median_diffs$diff <= -600) + mean(median_diffs$diff >= 600)
```


## Confidence intervals

Let's build a confidence interval for the average Adelie penguin flipper length.

```{r}
adelie %>%
  summarize(mean_flip_len = mean(flipper_length_mm))
```

```{r}
average_lengths <- tibble(
  run = 1:10000,
  xbars = replicate(10000,
                    mean(new_data(adelie, 146)$flipper_length_mm) ) 
)


average_lengths
```

Let's consider the 99% most central values (the 0.5th to 99.5th percentiles):

```{r}
quantile(average_lengths$xbars, c(0.005, 0.995))

m_lower <- quantile(average_lengths$xbars, 0.005)
m_upper <- quantile(average_lengths$xbars, 0.995)
```

The final CI calculation:

```{r}
m_obs <- 190.1027	

c(2*(m_obs) - m_upper, 2*m_obs - m_lower)
```

EX 7.3

Use the same data to build a 95% and a 98% CI.  You don't need to redo the sampling experiment - just get the correct quantiles for each confidence level!  Note how the width changes as the confidence level changes.

```{r}

```

EX 7.4

If time permits, build a 95% CI for median Gentoo penguin flipper length.  Here is the resampling experiment we did last time:

```{r}
gentoo <- penguins %>% filter(species == "Gentoo")

gentoo %>%
  summarize(med_length = median(flipper_length_mm))
```

```{r}
median_lengths <- tibble(
  run = 1:10000,
  medians = replicate(10000,
                    median(new_data(gentoo, 119)$flipper_length_mm) ) 
)

median_lengths
```