---
title: "Section 4: Introduction to ggplot2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

## Palmer penguins data

The Palmer penguin data records several physical measurements of three penguin species from Palmer Station, Antarctica.  Use the code below to load in a tidy version of this data, from bwu62.  The .csv file must be in your `data` folder.

```{r}
## This assumes that:
### STAT240/data/ contains the data file
### STAT240/lecture/sect04-ggplot/ is your working directory.
### If this gives you "Error: could not find file ... in working directory ...", go to Session > Set Working Directory > To Source File Location, and try again.
### If that doesn't work, then you downloaded one or both files to the wrong place, or they have the wrong name - make sure they don't have a " (1)" or "-1" at the end of their names, which can happen when you download multiple times.

penguins <- read_csv("../../data/penguins.csv")
```

Now explore the data.

```{r}
glimpse(penguins)
```


## Visual summaries

```{r}
# Basic empty canvas
ggplot()

# We defined the mapping - where is the plot?
ggplot(data = penguins,
       mapping = aes(x = flipper_length_mm))
```

To actually build the plot, we need to specify a geom and add it to the plot.

```{r}
ggplot(data = penguins,
       mapping = aes(x = flipper_length_mm)) +
  geom_histogram()
```

Let's add a fill aesthetic based on species.

```{r}
ggplot(data = penguins,
       mapping = aes(x = flipper_length_mm, fill = species)) +
  geom_histogram()
```

Now, let's change the bins.

```{r}
ggplot(data = penguins,
       mapping = aes(x = flipper_length_mm, fill = species)) +
  geom_histogram(binwidth = 5, boundary = 200)
```

Finally, let's add a title and better labels.

```{r}
ggplot(data = penguins,
       mapping = aes(x = flipper_length_mm, fill = species)) +
  geom_histogram(binwidth = 5, boundary = 200) +
  labs(
    title = "Histogram of Penguin Flipper Lengths",
    x = "Flipper Length (mm)",
    y = "Count",
    fill = "Species"
  )
```

Boxplots visualize the quartiles of our data, rather than the shape.

```{r}
ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_boxplot()

ggplot(penguins,
       aes(x = flipper_length_mm, fill = species)) +
  geom_boxplot()
```

### Plotting categorical data

Bar graphs depict a single categorical variable, like `species`.  `geom_bar()` counts the number of instances in a dataset.

```{r}
ggplot(penguins, aes(x = species)) +
  geom_bar()
```

With `geom_col()`, you specify the heights of the bars yourself.  It is more flexible than `geom_bar()` but more work to use.

Let's make a summarized version of the data with the counts and average flipper length of each species.

```{r}
# We'll cover this code next week!
penguin_summary <- penguins %>%
  group_by(species) %>%
  summarize(n_penguins = n(),
            mean_length = mean(flipper_length_mm))

penguin_summary
```

```{r}
# This code recreates the geom_bar graph
ggplot(penguin_summary, aes(x = species, n_penguins)) +
  geom_col()

# But we can make other plots as well
ggplot(penguin_summary, aes(x = species, mean_length)) +
  geom_col()
```

### Plotting two numeric variables

```{r}
ggplot(penguins,
       aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point()
```

We can use `geom_smooth` to draw a trendline.

```{r}
ggplot(penguins,
       aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point() +
  geom_smooth()

ggplot(penguins,
       aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point() +
  geom_smooth(method = "lm")
```

`geom_line` builds a line plot.

```{r}
# This does not really make sense
ggplot(penguins,
       aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_line()

# Enrollment data is a time series, so a line plot makes sense
enrollment <- read_csv("https://bwu62.github.io/stat240-revamp/data/enrollment.csv")

ggplot(enrollment, aes(x = year, y = enrolled_millions,
                       color = sex, linetype = sex)) +
  geom_line()
```

### Customization

Add customization options within the geom itself.  These options change the look of the graph but not the information.

```{r}
ggplot(penguins,
       aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point()

ggplot(penguins,
       aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(color = "blue", alpha = 0.5,
             size = 3, shape = 18)
```


Exercise 4.2:

```{r}
# Copy-paste one of the plots we made above and
# change one of the visual aspects shown in the plot.
```

## Numeric summaries

Let's find the mean and SD of the flipper length data.

```{r}
flip_len <- penguins$flipper_length_mm

# mean "long way"
n <- length(flip_len)
(1/n) * sum(flip_len)

# mean with R
mean(flip_len)
```

```{r}
# SD "long way"
xbar <- mean(flip_len)

sqrt( (1/(n-1)) * sum((flip_len - xbar)^2) )

# SD with R
sd(flip_len)
```

Let's add this as an annotation to our plot.

```{r}
xbar <- mean(flip_len)
s <- sd(flip_len)

##### We've seen this plot before
p <- ggplot(data = penguins,
       mapping = aes(x = flipper_length_mm, fill = species)) +
  geom_histogram(binwidth = 5, boundary = 200)
#####

useful_lines <- xbar + c(-2, -1, 0, 1, 2)*s

p +
  geom_vline(xintercept = useful_lines,
             linetype = c(2, 2, 1, 2, 2)) +
  labs(subtitle = "-2, -1, 0, 1, 2 SDs from Mean")

```

The median, range, and IQR are based on percentiles (quantiles) of the data.

```{r}
# Two ways of finding the median
median(flip_len)
quantile(flip_len, 0.5)

# What's the 30th percentile?  What about the 75th?
quantile(flip_len, c(0.3, 0.75))
```

```{r}
# IQR: 75th - 25th percentile
quantile(flip_len, c(0.25, 0.75))
IQR(flip_len)

# Range: min and max
range(flip_len)
```

A boxplot shows the quartiles (min, 25th, median, 75th, max) of the data.

```{r}
ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_boxplot()
```

Outlying points are drawn as dots.  An outlier is defined as 1.5 "box lengths" (1.5xIQR) away from the edges of the box.

Let's compare the mean and median flipper length.

```{r}
p +
  geom_vline(xintercept = mean(flip_len)) +
  geom_vline(xintercept = median(flip_len), lty = 3) +
  labs(subtitle = "Mean = Solid, Median = Dotted")
```



## Variable aesthetics

Let's look at a scatterplot of bill depth vs flipper length.

```{r}
ggplot(penguins, aes(x = bill_depth_mm, y = flipper_length_mm)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", se = F)
```

What if we want to color the points by species?

```{r}
ggplot(penguins, aes(x = bill_depth_mm, y = flipper_length_mm,
                     col = species)) +
  geom_point() +
  geom_smooth(method = "lm", se = F)
```

What if we want the points to be colored, but keep only a single smooth line?  We have to specify the color mapping within `geom_point()` rather than the original `ggplot()`.  This is called a *local* aesthetic.

```{r}
ggplot(penguins, aes(x = bill_depth_mm, y = flipper_length_mm)) +
  geom_point(aes(col = species)) +
  geom_smooth(method = "lm", se = F)
```

What's wrong with the following plots?  How can we fix them?

```{r}
# Read and interpret the error messages from these two plots
ggplot(penguins) +
  geom_point(aes(color = species))

ggplot(penguins, aes(x = bill_depth_mm, y = flipper_length_mm)) +
  geom_point(color = species)
```

```{r}
# Why are these points not huge? Why is there a legend for "size"?
ggplot(penguins, aes(x = bill_depth_mm, y = flipper_length_mm)) +
  geom_point(aes(color = species, size = 1000))

# This just produces a gridded canvas with no points. Why?
ggplot(penguins, aes(x = bill_depth_mm, y = flipper_length_mm,
                    color = species))
  geom_point()
```


## More customization

There are countless options for customization in ggplot2.  There is tons of information here - I encourage you to use this section as a reference that you can return to as needed.

### Text annotations

Use `annotate()` to add text annotations on top of the plot.  For example, label a vertical line for the mean.

```{r}
xbar <- mean(flip_len)

ggplot(penguins, aes(x = flipper_length_mm, fill = species)) +
  geom_histogram(binwidth = 5, boundary = 200) +
  geom_vline(xintercept = xbar,
             size = 1.5) +
  annotate("text", x = 220, y = 50,
                label = "Mean Flipper Length")
```

Or, add the values of the quartiles on top of a boxplot.

```{r}
quartiles <- quantile(flip_len, c(0.25, 0.5, 0.75))
  
ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_boxplot() +
  annotate("text", x = quartiles, y = 0.45,
                label = quartiles)
```


### Axis scales + labels

Scales allow us more fine-grained control over how aesthetics are displayed.  For example, we might want to customize the spacing of the x or y axes.

```{r}
ggplot(penguins, aes(x = flipper_length_mm)) + 
  geom_histogram(
    binwidth = 5,
    boundary = 70,
    color = "steelblue4",
    fill = "skyblue1"
  )

ggplot(penguins, aes(x = flipper_length_mm)) + 
  geom_histogram(
    binwidth = 5,
    boundary = 70,
    color = "steelblue4",
    fill = "skyblue1"
  ) +
  # Custom x axis
  scale_x_continuous(breaks = seq(170, 240, by = 10))
```

```{r}
ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_boxplot()

# Add some extra spacing to the y-axis in the boxplot
ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_boxplot() +
  scale_y_continuous(limits = c(-1, 1))
```

This also works with discrete scales.

```{r}
ggplot(penguins, aes(x = species, fill = sex)) +
  geom_bar(position = "dodge")

scientific_names <- c("Pygoscelis adeliae", "Pygoscelis antarcticus",
                      "Pygoscelis papua")

ggplot(penguins, aes(x = species, fill = sex)) +
  geom_bar(position = "dodge") +
  scale_x_discrete(label = scientific_names)
```

### Color palettes

We can also use custom color schemes for the `color` and `fill` aesthetics (also can be continuous or discrete). The default color scheme can be difficult to distinguish for certain types of color blindness.

```{r}
ggplot(penguins, aes(x = flipper_length_mm, fill = species)) +
  geom_density(alpha = 0.3)
```

We can use the pre-built viridis palettes for either the fill or the color aesthetic.

```{r}
ggplot(penguins, aes(x = flipper_length_mm, fill = species)) +
  geom_density(alpha = 0.3) +
  scale_fill_viridis_d()

ggplot(penguins, aes(x = flipper_length_mm, fill = species)) +
  geom_density(alpha = 0.3) +
  scale_fill_viridis_d(option = "inferno")
```

We can also make our own custom color scheme.

```{r}
ggplot(penguins, aes(x = flipper_length_mm, fill = species)) +
  geom_density(alpha = 0.3) +
  scale_fill_manual(
    values = c("Adelie" = "dodgerblue",
               "Chinstrap" = "peachpuff",
               "Gentoo" = "mediumorchid")
    )
```

Here's an example of a continuous color scheme.

```{r}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm,
                     color = flipper_length_mm)) +
  geom_point() +
  scale_color_viridis_c(option = "cividis")


ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm,
                     color = flipper_length_mm)) +
  geom_point() +
  scale_color_gradient(low = "tomato", high = "dodgerblue")
```



