---
title: "Section 5: Data Tidying"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

```{r}
library(tidyverse)
```


## The pipe

The pipe passes the object on the left to the function on the right.
(Ctrl + Shift + M) or (Command + Shift + M)

```{r}
sqrt(4)
4 %>% sqrt()  # sqrt(4) is 2

log(1000, base = 10)
1000 %>% log(base = 10)  # Log base 10 of 1000 is 3
```

The pipe allows for greater readability with a long chain of operations.  What if we need to do a long chain of operations?

To generate a user's email, add their last name initial to their first name, add a random digit, add the @ address, and make sure everything is lowercase.

```{r}
firstnames <- c("Lourens", "Mervyn", "Genevieve", "Petronella")
lastnames <- c("Eberly", "Stevenson", "Lombardt", "Purvess")

# Unintuitive / unreadable way
tolower(paste(substr(lastnames, 1, 1), firstnames,
                      sample(0:9, 4, replace=T),
                      "@stat240.com", sep = ""))

# Intuitive / readable way
lastnames %>%
  # Get 1st letter
  substr(1, 1) %>%
  # Add first name
  paste(firstnames, sep = "") %>%
  # Add random digit
  paste(sample(0:9, 4, replace = T), sep = "") %>%  
  # Add address
  paste("@stat240.com", sep = "") %>%
  # Make lowercase
  tolower()
```


## dplyr basics

We'll continue working with the Palmer penguins data as an example.

```{r}
penguins <- read_csv("../../data/penguins.csv")
```


### Column editors

`relocate()` moves columns without adding or removing any.  By default, it takes columns and moves them to the *left*.

```{r}
penguins %>% 
  relocate(year, sex)
```

Or, we can move columns to specific places.

```{r}
penguins %>% 
  relocate(island, .after = year)
```

`rename()` changes the column names.
Write `new_name = old_name`.

```{r}
penguins %>%
  rename(species_name = species, island_name = island)
```

`select()` reduces the df to the columns we specify.

```{r}
# select just species, sex, flipper length, and body mass
penguins %>%
  select(species, sex, flipper_length_mm, body_mass_g)

# Can also "de-select"
# e.g. select everything except island and
# columns body mass through year

penguins %>%
  select(-island, -(body_mass_g:year))
```


Note that none of these actions are changing the `penguins` object.  If we want to save the results, we can use `<-`, either to define a new object or overwrite the current object.

```{r}
# Let's make a smaller penguin dataset that's easier to see
penguins_small <- penguins %>%
    select(species, sex, bill_length_mm, bill_depth_mm)

penguins_small
```

`mutate()` creates a new column.  For example, make a new column whose value is constant for all rows.

```{r}
penguins_small %>%
  mutate(study = "Palmer")
```

Or, we can change the value of an existing column.

```{r}
penguins_small %>%
  mutate(sex = substr(toupper(sex), 1, 1))
```

More commonly, create a new column based on the values of other columns.  Let's estimate the volume of a penguin bill as a cylinder.

```{r}
penguins_small %>%
  mutate(bill_cross_section_mm2 = pi * (bill_depth_mm / 2)^2,
         
         bill_vol_mm3 = bill_cross_section_mm2 * bill_length_mm
)
```

By using `case_when()`, we can mutate a categorical column based on a condition.

```{r}
penguins_small %>%
  mutate(type = case_when(bill_length_mm < 40 ~ "small bill",
                          bill_length_mm < 48 ~ "med bill",
                          .default = "large bill")
         )
```


### Row editors

`arrange()` lets us order the rows while keeping the columns unchanged.  It sorts rows from smallest to largest.

```{r}
penguins %>% 
  arrange(flipper_length_mm)

penguins %>%
  arrange(island)
```

If we want to go from largest to smallest, use `desc`.

```{r}
penguins %>%
  arrange(desc(island))

# Advanced note that you don't have to know: R considers
# uppercase as "greater" than lowercase IF IT IS THE SAME LETTER.
# Otherwise, it ignores case.
# So, "B" > "b", but "A" < "b".
```

You can also provide further variables to break ties in the first one, if desired. If not, the order in the original dataframe will break the tie.


The `filter()` command chooses specific rows based on a condition.  In R, conditions evaluate to `TRUE` or `FALSE`.

```{r}
penguins

penguins$species 

# For each row in the data, check whether its species is Chinstrap
penguins$species == "Chinstrap" 
```

Let's keep only the rows corresponding to Chinstrap penguins.

```{r}
penguins %>% 
  filter(species == "Chinstrap")
```

We can set more complicated conditions with `&` (AND) and `|` (OR).

```{r}
# Only look at rows where body mass is between 3000 and 4000
penguins %>% 
  filter(body_mass_g >= 3000 & body_mass_g <= 4000)

# Look at rows where body mass is less than 3000 OR above 4000
penguins %>% 
  filter(body_mass_g < 3000 | body_mass_g > 4000)
```

We can use `slice_max()` or `slice_min()` to get the row(s) with the most extreme value in a given column.

```{r}
# Which row has the longest flipper length?
penguins %>%
  slice_max(flipper_length_mm) 

# Which one has the smallest body mass?
penguins %>%
  slice_min(body_mass_g)

# We can also ask for the smallest n
penguins %>%
  slice_min(body_mass_g, n = 5)
```

We can also set `with_ties` to be FALSE.  The function will return ties by default.

```{r}
# We can also ask for the smallest n
penguins %>%
  slice_min(body_mass_g, n = 5, with_ties = F)
```


`drop_na()` removes any rows with missing values.  In the `starwars` dataset, we have `NA` for the hair color of droids.

```{r}
# C-3PO and R2-D2 and R5-D4's hair_color are NA.
starwars

# Normally, there are 87 characters
nrow(starwars)
```

Now let's get rid of those rows.

```{r}
# Goodbye to those rows! Now we skip right from Luke to Darth Vader.
starwars %>% 
  drop_na(hair_color)

# After removing rows with NA, 82 rows are left
starwars %>% 
  drop_na(hair_color) %>% 
  nrow()
```

By default, `drop.na` looks for missing values in ALL columns.

```{r}
drop_na(starwars)
```

If we only want to drop NA values from a single column, specify the column name or use `filter`.

```{r}
starwars %>%
  drop_na(hair_color)

starwars %>% 
  filter(!is.na(hair_color))
```


Exercise 5.1

Friends A and B are fighting over who is the most die-hard Star Wars fan. Settle the debate with the built-in `starwars` data and use dplyr verbs to help answer their requests.

A: "The oldest character is an Ewok!"
B: "The oldest character is of Yoda's species!"

```{r}

```


A: "Kamino and Utapau are the only homeworlds with characters that have black eyes and are over 200 cm tall!"
B: "Kamino and Utapau are the only homeworlds with characters that have black eyes or are over 200 cm tall!"

```{r}

```


A: "The Droid characters appear in decreasing order of age!"
B: "No, because IG-88 is older than R5-D4!"

```{r}

```


A: "There's no character with a name longer than 20 letters!"
B: "The longest names have 21 letters!"

```{r}

```


A: "Female characters are the group with the shortest average height!"
B: "Female characters are only the group with the second shortest average height!  Characters with no biological sex have the shortest average height!"

```{r}

```


### summarize

`summarize()` reduces columns to a single number with a "reducing function".

Let's think about how R does operations with vectors.  Some operations return the same number of elements as the original vector.

```{r}
prime_numbers <- c(2, 3, 5, 7, 11)
fibonacci_numbers <- c(0, 1, 1, 2, 3)

prime_numbers + fibonacci_numbers

prime_numbers^2
```

Other operations reduce a vector to a single number.

```{r}
sum(prime_numbers)

max(fibonacci_numbers)

mean(prime_numbers)
```

When using `summarize()`, we apply a reducing function to a column (or columns) in our data set.

```{r}
penguins %>%
  summarize(avg_mass = mean(body_mass_g),
            min_mass = min(body_mass_g),
            max_mass = max(body_mass_g))
```

`summarize()` becomes even more powerful when combined with the `group_by()` command.

### grouop_by

`group_by()` is a powerful command that changes the *grouping* structure of a df.  On its own, it does not change any rows or columns:

```{r}
penguins %>% 
  group_by(species)
```

What if we want to find the average, min, and max body mass of the different species?  So far, we have a naive approach:

```{r}
penguins %>%
  filter(species == "Adelie") %>%
  summarize(avg_mass = mean(body_mass_g),
            min_mass = min(body_mass_g),
            max_mass = max(body_mass_g))

penguins %>%
  filter(species == "Chinstrap") %>%
  summarize(avg_mass = mean(body_mass_g),
            min_mass = min(body_mass_g),
            max_mass = max(body_mass_g))

# and so on for all of the species...
```

`group_by()` is a much more convenient way to perform the same operation multiple times.

```{r}
penguins %>%
  group_by(species) %>%
  summarize(avg_mass = mean(body_mass_g),
            min_mass = min(body_mass_g),
            max_mass = max(body_mass_g))
```

We can even group by multiple variables:

```{r}
penguins %>%
  group_by(species, sex) %>%
  summarize(avg_mass = mean(body_mass_g),
            min_mass = min(body_mass_g),
            max_mass = max(body_mass_g))
```


We use `mutate` to create a new column, usually based on the values of other columns.

```{r}
penguins %>% 
  mutate(new_column = 5) %>% 
  relocate(new_column)

penguins %>% 
  mutate(avg_mass = mean(body_mass_g)) %>% 
  relocate(avg_mass)
```

When we use `group_by()`, each row gets assigned a number based on group.  For example, we find average body mass separately by species.

```{r}
penguins %>% 
  group_by(species) %>% 
  mutate(avg_mass = mean(body_mass_g)) %>% 
  relocate(avg_mass)
```

What is the standard score for each penguin's body mass?

```{r}
penguins %>% 
  mutate(avg_mass = mean(body_mass_g),
         sd_mass = sd(body_mass_g),
         
         std_mass = (body_mass_g - avg_mass) / sd_mass) %>% 
  
  relocate(std_mass)
```

Now, let's try grouping by species first.

```{r}
penguins %>% 
  group_by(species) %>% 
  mutate(avg_mass = mean(body_mass_g),
         sd_mass = sd(body_mass_g),
         
         std_mass = (body_mass_g - avg_mass) / sd_mass) %>% 
  
  relocate(std_mass)
```


Finally, use `ungroup()` to remove the grouping property.

```{r}
largest <- penguins %>%
  group_by(species) %>%
  slice_max(body_mass_g, n = 5)

largest

largest %>%
#  ungroup() %>%
  summarize(avg_mass = mean(body_mass_g))
```


Exercise 5.2: BEST PENGUIN BEAK CONTEST!

Suppose that the volume of a Chinstrap or Gentoo penguin's beak is approximated by $$(\pi\times \text{radius}^2 \times length) / 2.5$$, and the volume of an Adelie penguin's beak is approximated by $$(\pi\times \text{radius}^2 \times length) / 2$$.

- Calculate each penguin's approximate beak volume (pay attention to radius vs depth!)
- Find the penguin with the largest beak volume for each combination of species and island.
- Were most of the winners measured in 2007 or 2009?

```{r}

```

## Joining

The dataset `grocery-list.csv` shows a customer's grocery list, while `supermarket.csv` shows the type and price of produce items at the supermarket.

```{r}
grocery_list <- read_csv("../../data/grocery-list.csv")

supermarket <- read_csv("../../data/supermarket.csv")

grocery_list
supermarket
```

Let's combine this information into a single dataset.


## Mutating joins

`left_join()` will merge the datasets and keep all values in the leftmost (x) dataframe.  We specify the column to join by.

```{r}
left_join(grocery_list, supermarket)

# Pipe notation
grocery_list %>% 
  left_join(supermarket)

# Now we can calculate the total cost of the customer's items
grocery_list %>% 
  left_join(supermarket) %>%
  mutate(cost = quantity * price)
```

`right_join()` does the opposite, by keeping all values in the rightmost (y) dataframe.  Its functionality is identical to `left_join()`.

```{r}
right_join(grocery_list, supermarket)

# Same result (but different order of rows+columns):
left_join(supermarket, grocery_list)
```

`left_join(x, y)` is identical to `right_join(y, x)`.


### Full join

`full_join()` will keep all rows in both x and y.  Imagine doing a right or left join, and just appending any extra, non-matching rows at the bottom.

```{r}
full_join(grocery_list, supermarket)
```

This is different from simply stacking the rows.

```{r}
bind_rows(grocery_list, supermarket)
```


### Inner join

`inner_join()` is the most "selective".  It only keeps rows that appear in both datasets and it does not induce any NA values.

```{r}
inner_join(grocery_list, supermarket)
```

### Filtering joins

Filtering joins decide to keep or remove rows, similar to `filter()`.  `semi_join()` returns matching rows.

What items on the list can be found at the store?  The order matters here.

```{r}
grocery_list %>% 
  semi_join(supermarket)

supermarket %>% 
  semi_join(grocery_list)
```

Notice the differences with `inner_join()`.  `inner_join()` results in a new column and `semi_join()` does not.

```{r}
inner_join(grocery_list, supermarket)
```


Use `anti_join()` to filter by non-matching rows.  What items on the list are not at the store?

```{r}
grocery_list %>% 
  anti_join(supermarket)
```

What items at the store are not on my grocery list?

```{r}
supermarket %>% 
  anti_join(grocery_list)
```


### Different column names

How does R know how to join the dataframes?  In the supermarket example, it's easy, since both dfs have a column called `item`.

We can specify this ourselves with the `by` argument.

```{r}
left_join(grocery_list, supermarket)
```


Let's look at `join_by()` in more detail. In the original discounts and drinks data, both have a column named `item`, which R can recognize.

```{r}
left_join(grocery_list, supermarket,
          by = join_by(item))

# More specifically: use == to connect both
# dataframe column names
left_join(grocery_list, supermarket,
          by = join_by(item == item))
```

Exercise 5.3:

The `drinks` dataframe contains information on a user's favorite cafes and the drinks they get at each cafe.  It also has the price of each drink and the date the user last visited.

The `coupons` dataframe contains coupons from several cafes with the corresponding discount (e.g. 0.25 means 25% off) and the expiration date.

- Join the dataframes together with a mutating join such that the final dataframe only has information on the user's favorite drinks.  Make sure the correct columns are joined!

- Mutate an additional column for each drink's discounted price.  If there's no discount, the discounted price should be the normal price.

```{r}
drinks <- tibble(
  Date = c("8/15", "9/7", "9/7", "8/22"),
  Drink = c("Milk Tea", "Espresso", "Latte", "Espresso"),
  Cafe = c("Happy Tea", "Daily Grind", "Daily Grind", "Bev's Beverages"),
  Price = c(6, 4, 5, 5)
)

coupons <- tibble(
  Discount = c(0.25, 0.5, 0.05),
  Date = c("9/30", "10/31", "10/15"),
  Location = c("Happy Tea", "Daily Grind", "The Roast")
)


```























