---
title: "Section 6: Simulation and Resampling"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

```{r}
library(tidyverse)
```


## Probability simulations

Let's simulate 10000 rolls of a fair six-sided die.

```{r}
dice_example <- tibble(rolls = sample(1:6,
                                      size = 10000,
                                      replace = T))

dice_example
```

The barplot of counts looks roughly flat, and the simulated average is close to the theoretical average.

```{r}
ggplot(dice_example, aes(x = rolls)) +
  geom_bar(fill = "goldenrod2", color = "black")
```

```{r}
mean(dice_example$rolls)
```


What about something that's difficult or annoying to study analytically?

The custom function below, `generate_score()`, randomly generates a strength score for a D&D character by rolling four six-sided dice and taking the sum of the highest three.

```{r}
generate_score <- function() {
  rolls <- sample(1:6, 4, replace = TRUE)   # Roll dice
  keep <- sort(rolls)[2:4]       # Keep highest 3 rolls
  sum(keep)
}
```

This function is random, and it will return a different result each time.  

```{r}
generate_score()
```

By sampling many, many, times, we can create an empirical/simulated histogram of possible strength scores.

```{r}
strength_values <- tibble(
  run = 1:10000,
  values = replicate(10000, generate_score())
)

ggplot(strength_values, aes(x = values)) +
  geom_histogram(binwidth = 1, boundary = 2.5,
                 fill = "goldenrod2", color = "black")
```

We can also get a mean, median, standard deviation, and IQR from the simulated values.

```{r}
strength_values %>%
  summarize(avg_strength = mean(values),
            med_strength = median(values),
            sd_strength = sd(values),
            iqr_strength = IQR(values))
```

Now, consider generating all six scores for a character.

```{r}
all_character_scores <- function(){
  setNames(replicate(n = 6, generate_score()),
           c("STR", "DEX", "CON", "INT", "WIS", "CHA"))
}
```

```{r}
my_stats <- all_character_scores()

my_stats
```

What is this character's average score?

```{r}
mean(my_stats)
```

EX 6.1

```{r}
# Some code to get you started.
# What process are we repeating?

average_scores <- tibble(
  run = 1:10000,
  values = replicate(10000, mean(all_character_scores())) 
)

average_scores
```

```{r}
ggplot(average_scores, aes(x = values)) +
  geom_histogram(fill = "dodgerblue", color = "black")
```



In statistics (unlike D&D scores), the process to generate data is unknown.  Let's consider the population of flipper lengths for all Adelie penguins.

We don't have this population - only a sample of size 146.

```{r}
penguins <- read_csv("../../data/penguins.csv") %>%
  relocate(flipper_length_mm, .after = species)

adelie <- penguins %>% filter(species == "Adelie")
```

This data has a sample mean of 190.1 mm.  But is this a good representation of the *true* mean?

```{r}
adelie %>%
  summarize(mean_flip_len = mean(flipper_length_mm))
```

How can we learn about the population via resampling, if we only have 146 observations?

The data represents the population.  So, by resampling from this data with `sample_n`, we are approximating a sample from the full population.

```{r}
sample_n(adelie, size = 146, replace = TRUE)
```


```{r}
ggplot(adelie, aes(x = flipper_length_mm)) +
  geom_histogram()
```


Each time we resample from the data, we calculate a new mean flipper length.

```{r}
resample <- sample_n(adelie, size = 146, replace = TRUE)

resample %>%
  summarize(mean_flip_len = mean(flipper_length_mm))
```

Here's the same code, wrapped in a function.

```{r}
new_data <- function(data, n) {
  sample_n(data, size = n, replace = TRUE)
}

new_data(adelie, 146)
```

Now, let's replicate this to simulate 10000 different penguin studies.

```{r}
average_lengths <- tibble(
  run = 1:10000,
  xbars = replicate(10000,
                    mean(new_data(adelie, 146)$flipper_length_mm) ) 
)


average_lengths
```

We can calculate graphical and numeric summaries of these simulated x-bars. 

```{r}
ggplot(average_lengths, aes(x = xbars)) +
  geom_histogram(fill = "dodgerblue", color = "black")
```

```{r}
adelie %>%
  summarise(avg_length = mean(flipper_length_mm),
            sd_length = sd(flipper_length_mm))

average_lengths %>%
  summarize(xbar_mean = mean(xbars),
            xbar_sd = sd(xbars),
            xbar_IQR = IQR(xbars))
```

We can also compare this to the original flipper length data.

```{r}
adelie %>%
  summarize(flip_mean = mean(flipper_length_mm),
            flip_sd = sd(flipper_length_mm),
            flip_IQR = IQR(flipper_length_mm))
```


EX 6.3

Now, repeat the resampling, but study the median of Gentoo penguin flipper lengths instead of the mean of Adelie penguin flipper lengths.  You can copy/paste most of the code from above and use the new_data() function from before.

```{r}
gentoo <- penguins %>% filter(species == "Gentoo")

gentoo %>%
  summarize(med_length = median(flipper_length_mm))
```

```{r}
median_lengths <- tibble(
  run = 1:10000,
  medians = replicate(10000,
                    median(new_data(gentoo, 119)$flipper_length_mm) ) 
)

median_lengths
```

```{r}
ggplot(median_lengths, aes(x = medians)) +
  geom_histogram(fill = "dodgerblue", color = "black")
```


## Resampling two populations

Now, consider comparing the median body mass for male vs female Adelie penguins.

```{r}
adelie %>%
  group_by(sex) %>%
  summarize(mass_med = median(body_mass_g), n = n())
```

If we assume the median body mass is the same for both groups, we can treat the entire population of male+female penguins as a "big box" for resampling.

```{r}
new_female_med <- median(new_data(adelie, 73)$body_mass_g)
new_male_med <- median(new_data(adelie, 73)$body_mass_g)

new_female_med - new_male_med
```

Let's repeat this 10000 times.

```{r}
median_diffs <- tibble(
  run = 1:10000,
  female = replicate(10000,
                    median(new_data(adelie, 73)$body_mass_g) ),
  male = replicate(10000,
                    median(new_data(adelie, 73)$body_mass_g) ),
  diff = female - male
)

median_diffs
```

```{r}
ggplot(median_diffs, aes(x = diff)) +
  geom_histogram(fill = "dodgerblue", color = "black")
```

The actual, observed difference in medians (female - male) is -600.  Let's compare that to our histogram.

```{r}
ggplot(median_diffs, aes(x = diff)) +
  geom_histogram(fill = "dodgerblue", color = "black") +
  geom_vline(xintercept = -600,
             color = "forestgreen", linewidth = 1.5)
```


