---
title: 'Section 3: Data Types and Structures'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

## R basics

```{r}
# Initialize with <- or =, but <- is standard R style
x <- 4
my_favorite_number <- 8

# This does not change the value of either x or my_favorite_number
x + my_favorite_number

# But if we do want to update the value:
x <- x + 1
```

Capitalization matters!

```{r}
x

X
```

```{r}
# This is an illegal variable name
1a <- 3
```


Three variable types: numeric, character, and logical.  We're familiar with numerics as a way of storing numbers.

```{r}
class(my_favorite_number)

class(pi)
```

Math operations can result in `NaN`, `Inf` or `-Inf` (to avoid breaking R).

```{r}
sqrt(-4)
0 / 0

1 / 0

log(0)
```


Character objects are a set of symbols like a name or label.  They have no mathematical or other special meaning.

```{r}
my_name <- "Miranda"

class(my_name)
```

Characters don't obey numeric or logical operators.

```{r}
easy_equation <- "2 + 2"
false_statement <- "5 < 2"

easy_equation
false_statement

"2026" + 10
```

```{r}
nchar(my_name)

tolower(my_name)
toupper(my_name)
```

Use `paste()` to combine strings.

```{r}
my_name <- "Miranda"
my_age <- "29"

paste(my_name, "is", my_age, "years old.")
```


The two basic types of logical objects are `TRUE` and `FALSE`.

```{r}
# Is three greater than two? Yes, so return TRUE.
3 > 2

# Is two greater than three? No, so return FALSE.
2 > 3
```


## Logical operators

Mathematical operators like `+` let us compute a numeric result with numeric objects. Logical operators perform a comparison.

```{r}
my_height <- 60

my_height < 72

my_height < 60
```

```{r}
my_height <= 72
my_height >= 60
```

Use `==` and `!=` to check an exact equality.  This is not the same as the `=` for assigning variables!

```{r}
my_height == 66
my_height != 66
```

The value of `my_height` does not change.

The operators `&` and `|` also perform a comparison.  Instead of comparing numbers, they compare logical statements.

```{r}
# one of these fails, so the whole "and" condition fails.
(my_height > 72) & (my_height + 6 < 70)

# one of these succeeds, so the whole "or" condition succeeds.
(my_height > 72) | (my_height + 6 < 70)
```


```{r}
TRUE & FALSE
TRUE & TRUE

TRUE | TRUE
TRUE | FALSE
FALSE | FALSE
```


`NA` is the notation for a missing value.  It doesn't work with normal operators.

```{r}
nothing <- NA

nothing + 2
nothing > 5
nothing | TRUE
```

To check for a missing value, use `is.na`.

```{r}
# Incorrect (R can't interpret this statement):
nothing == NA

# Correct
is.na(nothing)
```


## Vectors

A vector is a sequence of the same type of object.  Access an item by index with `[]`.  Vectors in R are *one-indexed*.

```{r}
favorite_foods <- c("noodles", "lamb stew", "corn on the cob")
prime_numbers <- c(2, 3, 5, 7, 11)

# Notice how these variables are storing the entire sequence
favorite_foods
prime_numbers

favorite_foods[2]
prime_numbers[1]
```

If you try to combine different types, R will try to force the objects to be the same type.

```{r}
three_things = c(2, TRUE, "banana")

three_things

class(three_things)
```

We often want to make a numeric sequence following a specific pattern.  Use `:` to get consecutive whole numbers and `seq` to get a sequence.

```{r}
1:10

# Example use: setting axis breaks in a graph
seq(from = 0, to = 50, by = 5)
```

Most functions in R are *vectorized*. The operation is performed to the first element of each, then the second element of each, and so on.  We do not need to make a loop to do this type of operation.

```{r}
prime_numbers

prime_numbers + 13

# Use c() to add a new element
c(prime_numbers, 13)
```

This also works for logical operators.

```{r}
prime_numbers == 2

sum(prime_numbers == 2)
```

We can also use `%in%` to look for a specific item.

```{r}
2 %in% prime_numbers
1 %in% prime_numbers
```

Note that %in% is not vectorized: it only has a single output.

We can also do element-wise operations over two vectors.

```{r}
onetwothree <- 1:3
fourfivesix <- 4:6

onetwothree + fourfivesix

fourfivesix - onetwothree

onetwothree * fourfivesix
```

We have to be careful if the vectors are different sizes!

```{r}
fourfive <- 4:5

onetwothree + fourfive
```

Which of the following functions are vectorized?

```{r}
prime_numbers <- c(2, 3, 5, 7, 11)

log(prime_numbers)

min(prime_numbers)
max(prime_numbers)

sqrt(prime_numbers)

mean(prime_numbers)
sum(prime_numbers)

median(prime_numbers)
abs(prime_numbers)
```

Exercise 3.2:

Convert the course names to uppercase, e.g. "STAT240".

```{r}
c(toupper("Stat240"), toupper("cs300"), toupper("STAT309"))
```


For what values of $x$ is the following function negative?

$$(\frac{1}{2}x - 8)^2 - 20$$

```{r}
# The function has a negative value when x = 9
x <- 9

((1/2)*x - 8)^2 - 20
```

Re-write the following code so that it doesn't need to use a `for` loop.

```{r}
meals <- c("enchiladas", "stir fry", "baked potatoes",
           "pork chops", "casserole")

plan <- rep("", 5)

for (i in 1:5) {
    plan[i] <- paste("Meal", i, "is", meals[i])
}

plan

```


We can also use the `replicate()` command to repeat an operation $n$ times.

```{r}
# Works like `rep()`
replicate(n=5, FALSE)

# Roll a die 12 times
replicate(12, sample(1:6, 1))



```

```{r}
# Custom expression: make a new name of length l
new_word <- function(l) {
  paste0(sample(letters, size = l, replace = T), collapse = "")
}

new_word(8)

replicate(n = 9, new_word(5))
```

The `replicate()` command will be useful when we look at resampling-based methods.  It can also be used to generate random data for an experiment.


## Dataframes

Dataframes are a way to organize multiple vectors.  They're the usual tool for representing real life data.  We can build them with the `tibble` function from tidyverse.

```{r}
# "letters" is a built in R vector with the 26 lowercase letters
letters

library(tidyverse) # Need to do this to get tibble() to run

alphabet <- tibble(char = letters,
                   pos = 1:26,
                   is_vowel = letters %in% c("a", "e", "i", "o", "u"))
alphabet
```

Each row represents an individual letter.  We can subset a dataframe by index with the syntax `[,]`.  We can find a row by its name with `$`.

```{r}
# Value of 7th row, 1st column
alphabet[7, 1]

# Ninth row only
alphabet[9, ]

# Third column only
alphabet[ , 3]

alphabet$is_vowel
```


Below is an example of how to read in data from a url.  Make sure you have the tidyverse library loaded.

```{r}
eruptions_recent <- read_csv(
  "https://bwu62.github.io/stat240-revamp/data/eruptions_recent.csv",
  show_col_types = FALSE
)

# This opens a new tab in Rstudio
View(eruptions_recent)

# This gives a preview+summary of the data
glimpse(eruptions_recent)
```









